---
title: "Safegraph - Census data aggregation"
output: html_notebook
---

```{r}
# install packages
library(tidycensus)
library(tidyverse)
library(tigris)

# set trigris options
options(tigris_class="sf")
options(tigris_use_cache = T)
```

```{r}
# replace my census API key with your own census API key
# instructions to download:
census_api_key("c9dfbfcf8b82d02d96deb6d4a138f8e5e6e54050")
```

# Census data

## Load and clean ACS variables (race, edu, income)
```{r}
# for table names: https://censusreporter.org/topics/
# change from MD for other states

## MEDIAN INCOME
balt_med_inc = 83160

# get income for all tracts in STATE
balt_tract_inc = get_acs(geography = "tract",
                         variables = "B19013_001",
                         state = "MD",
                         year = 2019,
                         geometry = FALSE,
                         output = "tidy")

# create above/below median income variable
balt_tract_inc = balt_tract_inc %>%
  separate(NAME, c("tract","county","state"), sep = ",") %>%
  separate(tract, c(NA,NA,"tract_num"), sep = " ") %>%
  mutate(median_hhincome = estimate) %>%
  mutate(above_median_income = ifelse(median_hhincome > balt_med_inc, 1, 0))

# note -- can be same tract_num with different GEOID
# balt_tract_inc %>% filter(tract_num == 1)

## RACE

# create a list of census race variables (within summary table)
race_vars = c(
  White = "B03002_003",
  Black = "B03002_004",
  Hispanic = "B03002_012"
)

# get race variables (and summary table) for STATE
balt_tract_race = get_acs(geography = "tract",
                          state = "MD",
                          variables = race_vars,
                          summary_var = "B03002_001",
                          year = 2019,
                          geometry = FALSE,
                          output = "tidy")

# format race variables -- create pct vars and then group by geoid to eliminate redundant rows
balt_tract_race = balt_tract_race %>%
  separate(NAME, c("tract","county","state"), sep = ",") %>%
  separate(tract, c(NA,NA,"tract_num"), sep = " ") %>%
  spread(variable, estimate, fill=FALSE) %>%
  mutate(pct_white = White/summary_est * 100) %>%
  mutate(pct_black = Black/summary_est * 100) %>%
  mutate(pct_hispanic = Hispanic/summary_est * 100) %>%
  select(-White, -Black, -Hispanic) %>%
  group_by(GEOID) %>%
  summarize(pct_white = sum(pct_white), pct_black = sum(pct_black), pct_hispanic = sum(pct_hispanic))

balt_tract_race


## EDUCATION

# create a list of census edu variables (within summary table)
# https://www.socialexplorer.com/data/ACS2016_5yr/metadata/?ds=ACS16_5yr&table=B15003
# Less than high school graduate
# High school graduate (includes equivalency)
# Some college or associate's degree
# Bachelor's degree or higher 
edu_vars = c(
  no_school = "B15003_002",
  nursery_school = "B15003_003",
  K = "B15003_004",
  G1 = "B15003_005",
  G2 = "B15003_006",
  G3 = "B15003_007",
  G4 = "B15003_008",
  G5 = "B15003_009",
  G6 = "B15003_010",
  G7 = "B15003_011",
  G8 = "B15003_012",
  G9 = "B15003_013",
  G10 = "B15003_014",
  G11 = "B15003_015",
  G12_no_dip = "B15003_016",
  hs_dip = "B15003_017",
  ged = "B15003_018",
  some_col1 = "B15003_019",
  some_col2 = "B15003_020",
  assoc = "B15003_021",
  ba = "B15003_022",
  ma = "B15003_023",
  p_ma = "B15003_024",
  phd = "B15003_025"
)


balt_tract_edu = get_acs(geography = "tract",
                         state = "MD",
                         variables = edu_vars,
                         summary_var = "B15003_001",
                         year = 2019,
                         geometry = FALSE,
                         output = "tidy")

# create edu categories and make pcts
balt_tract_edu = balt_tract_edu %>%
  separate(NAME, c("tract","county","state"), sep = ",") %>%
  separate(tract, c(NA,NA,"tract_num"), sep = " ") %>%
  spread(variable, estimate, fill=FALSE) %>% 
  group_by(GEOID) %>%
  mutate(pct_less_than_hs = ((sum(c_across(cols = c("no_school","nursery_school","K","G1","G2","G3","G4",
                              "G5","G6","G7","G8","G9","G10","G11","G12_no_dip"))))/summary_est)*100) %>%
  mutate(pct_hs = ((sum(c_across(cols = c("hs_dip","ged"))))/summary_est)*100) %>%
  mutate(pct_some_college = ((sum(c_across(cols = c("some_col1","some_col2"))))/summary_est)*100) %>%
  mutate(pct_BAplus = ((sum(c_across(cols = c("ba","ma","p_ma","phd"))))/summary_est)*100) %>%
  select(GEOID,tract_num,pct_less_than_hs,pct_hs,pct_some_college,pct_BAplus)

balt_tract_edu = balt_tract_edu %>% distinct()


balt_tract_edu
```
```{r}
## join all census tract data together, select important columns
merged_census = merge(balt_tract_inc, balt_tract_race, by="GEOID") %>%
  merge(balt_tract_edu, by="GEOID") %>%
  select(GEOID,tract_num.x,above_median_income,pct_white,pct_black,pct_hispanic,pct_less_than_hs,
         pct_hs,pct_some_college,pct_BAplus) %>%
  rename(tract_num = tract_num.x)

merged_census
```


# Aggregate Safegraph block groups to tract level using Census data

## Get block groups for each GEOID string (cbg in safegraph)
```{r}
# Use income as a dummy variable to get Census block geography
blocks = get_acs(geography = "block group",
                 variables = "B19013_001",
                 state = "MD",
                 year = 2019,
                 geometry = TRUE,
                 output = "tidy"
)

# clean up columns to create tract_number ID, filter out all columns except for GEOID (same as cbg) and tract for merge
blocks_parsed = blocks %>%
  separate(NAME, c("block","tract","county","state"), sep = ",") %>%
  separate(tract, c(NA,NA,NA,"tract_num"), sep = " ") %>%
  separate(block, c(NA,NA,"block_num"), sep = " ") %>%
  mutate(tract_block = substring(GEOID,6)) %>%
  select(GEOID,tract_num)

```

## Read in Safegraph data and merge to get tract number for each cbg
```{r}
# read safegraph data -- change file for each MSA
msa = read.csv("/Users/esrieves/Documents/school/Research/foot_traffic/data/Outputs/normalized_output_data/Baltimore_MSA_18to21_visitor_flows/part-00000-f2026452-40bb-4ef7-8ba2-b9abfa636ac5-c000.csv")

# change sender_cbg to be called GEOID to facilitate merge with census data
msa = msa %>%
  rename(GEOID = sender_cbg,
         # named wrong in the other script, changing to reduce confusion
         monthly_visitors_per_naics_cbg_NORMALIZED = monthly_visitors_per_naics_tracts_NORMALIZED)

msa
```
```{r}
summary(msa)

msa %>% 
  group_by(GEOID) %>%
  summarise(n_cbg = n())

msa %>%
  group_by(GEOID) %>%
  summarise_all(~sum(is.na(.)))

# some CBGs are from other states (like Alabama).. I think that this might somehow lead to NAs for these same CBGs later.

# Also it points to the fact that Census data should be pulled in for all states in case of this same issue.
```


```{r}
# merge MSA Census data (dummy) with Safegraph to get the tract number for each block group
MSA_census = merge(msa,blocks_parsed, by="GEOID")

# select necessary columns and aggregate visitors from block to tract level
# keep geoid for merge with income data
MSA_census = MSA_census %>%
  select(date_start, naics, monthly_visitors_per_naics_cbg_NORMALIZED,tract_num,GEOID) %>%
  group_by(tract_num,date_start,naics) %>%
  summarise(visitors = sum(monthly_visitors_per_naics_cbg_NORMALIZED),GEOID = GEOID)

# preview data
MSA_census

# check to make sure that tract number makes sense
# while these tract numbers seemed strange, they are listed in the "blocks" df above, indicating that they are valid tract names
MSA_census %>% distinct(tract_num)
```

## Determine if tracts are in MSA
```{r}
# code to determine if tract is in a MSA

# Retrieve census tracts in STATE with the selected variable (total pop in this case)
# variable = dummy variable

# replace "md" with other state codes as needed
md_tracts = get_acs(geography = "tract", year = 2019, variables = "B01001_001", state = "md",
                    geometry = TRUE)

# replace "Baltimore" for other MSAs
msa = core_based_statistical_areas(cb = T) %>%
  filter(str_detect(NAME, "Baltimore"))

# retrieve census tracts in STATE that fall within the MSA
msa_tracts = md_tracts[msa,]

# make MSA tracts into a list with just tract names
msa_tract_list = msa_tracts %>%
  separate(NAME, c("tract","county","state"), sep = ",") %>%
  separate(tract, c(NA,NA,"tract_num"), sep = " ") %>%
  select(GEOID) %>%
  as.list()

#msa_tract_list
```

```{r}
# create a df of safegraph tracts (distinct removes duplicates)
safegraph_tracts = MSA_census$GEOID %>% as.data.frame() %>% distinct() %>% rename(GEOID = ".")

# creates is_in_MSA column 
safegraph_tracts = safegraph_tracts %>%
  mutate(is_in_MSA = ifelse(unlist(safegraph_tracts) %in% unlist(msa_tract_list),1,0))

# preview data
safegraph_tracts

# join census MSA with "is_in_MSA" data
MSA_census_joined = merge(MSA_census,safegraph_tracts,by="GEOID")


MSA_census_joined %>% 
  group_by(is_in_MSA) %>%
  summarize(n_MSA = n())
```

## Add Census vars to Safegraph data
```{r}
## ADD INCOME
# drop the block number digit from GEOID column to match up with census tract variables
MSA_census_joined$GEOID = substr(MSA_census_joined$GEOID,1,11)

# merge MSA data with Census income data
full_MSA = merge(MSA_census_joined, merged_census, by="GEOID")

# clean full_MSA dataset
full_MSA = full_MSA %>%
  select(-tract_num.y) %>%
  rename(tract_num = tract_num.x) %>%
  group_by(GEOID,date_start,naics)


full_MSA
```
# Descriptive stats of Safegraph-Census dataset
```{r}
summary(full_MSA)

# NAs for NAICS could be due to empty CBG phenomenon noticed above
# Census variable NAs likely related to CBG/tract out of state discovered above
```


# Transform to .R script
```{r}
library(knitr)

purl("safegraph_census_agg.Rmd")
```


